Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-12-18 manu-draft.qmd
Minor methods text revision suggestions

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b). These data are used for regulatory assessments for the nutrient TMDL in Tampa Bay. Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Time of sampling can vary, although most samples are collected from mid-morning to early afternoon. All surface and bottom salinity and temperature measurements for each of the 45 monitoring stations were evaluated herein. Trends were assessed for both surface and bottom samples, as described below, whereas only the bottom measurements were used for comparison to seagrass trends. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

The second dataset used to evaluate water quality trends was available from the Florida Fish and Wildlife Conservation Commission (FWC). The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple survey gear types that target different habitats. A stratified sample design is used to target multiple habitats at different sites sampled each month. We used data from the 21.3 meter center-bag nearshore seine that targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c). These data are collected by pulling the seine adjacent to the shore to sample approximately 140 m$^2$ of bay bottom [@schrandt2021]. In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the beginning and end of the seine haul, and typically at the surface and bottom. Only measurements at the beginning of the seine haul and from the bottom were used. Seagrass data are also provided for each site, with information on species and cover. Total percent cover for all species at a site was used for comparison with temperature and salinity measurements. Sites exclusively with macroalgae were not included in the analysis. All FIM data were provided from FWC staff by request.

The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM). Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting. Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>). We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments. Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone. Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained for analysis. Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.

All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program. This *ad hoc* group has routinely met quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data. Split-samples evaluated by each organization are also compared to assess precision between different laboratories. As such, the water quality measurements used herein are considered comparable, relative to the different sampling designs used by each program.

@@

Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was driven by climate stressors. The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (`r epcmodsum1`). An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab). For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016. However, the interaction between time period and temperature was not significant, suggesting no difference between time periods. The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure \[-@fig-sgmod\]a, @tbl-modtab). The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively. The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab). Note that Figure [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down (i.e., no interactions of bay segments with the other predictors were in the original models).

>>>>

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b). These data are used for regulatory assessments for the nutrient TMDL <!--# spell out --> in Tampa Bay. Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Time of sampling can vary, although most samples are collected from mid-morning to early afternoon. All surface and bottom salinity and temperature measurements for each of the 45 monitoring stations were evaluated herein. Trends were assessed for both surface and bottom samples, as described below, whereas only the bottom measurements were used for comparison to seagrass trends. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

The second dataset used to evaluate water quality trends was available from the Florida Fish and Wildlife Conservation Commission (FWC). The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple gear types that target different habitats. A stratified sampling design is used to select sites for 21.3 meter center-bag <!--# Looking at the code - you pulled Gear 20, which would include both shoreline and offshore (>5m from shore) sets; if you want just offshore sets select by strata A and B only. If both deployments kept, suggest deleting nearshore --> seines that target shallow habitats (\<1.5 m) where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c). Seines were deployed (1) parallel to bay shorelines and (2) offshore into the prevailing current at least 5 m from the shoreline to sample approximately 140 m$^2$ of bay bottom [@schrandt2021]. In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the bag, and at the surface and at 1 m intervals to the bottom. Only measurements from the bottom were used. Seagrass data are also provided for each site, with information on species and cover. Total percent cover for all species at a site was used for comparison with temperature and salinity measurements. Sites exclusively with macroalgae were not included in the analysis. All FIM data were provided from FWC staff by request.

The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM). Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES <!--# spell out --> reporting. Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>). We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments. Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone. Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained for analysis. Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.

All of the organizations that provided water quality datasets participate in the Southwest Florida Regional Ambient Monitoring Program <!--# Does FIM participate in this? -->. This *ad hoc* group has routinely met quarterly to ensure similar standards and methods are used for the collection and processing of surface water quality monitoring data. Split-samples evaluated by each organization are also compared to assess precision between different laboratories. As such, the water quality measurements used herein are considered comparable, relative to the different sampling designs used by each program.

@@

Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was driven by climate stressors. The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (`r epcmodsum1`). An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab). For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016. However, the interaction between time period and temperature was not significant, suggesting no difference between time periods. The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab). The interaction between the temperature and salinity metric was included in the most parsimonious model (marginally significant, p \< 0.1), although the shape of the relationship of temperature with time period or salinity with time period did not noticeably change using different conditional values for salinity or temperature, respectively. The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab). Note that Figure [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down (i.e., no interactions of bay segments with the other predictors were in the original models).
